---
title: "AquaX and IUCN status"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(iucnredlist) ### remotes::install_github("IUCN-UK/iucnredlist")
library(janitor)
```

This document finds the IUCN threat information for the AquaX species

```{r}
aquax_iucn <- read_csv("data/2024_ohara_iucn_to_worms_match.csv")
```

```{r}
api_key <- Sys.getenv("IUCN_API_KEY") ### usethis::edit_r_environ() to add key to R environment file
api <- init_api(api_key)
```

```{r}
# Get sample
aquax_iucn_30 <- aquax_iucn[1:30,]
```

```{r}
iucn_id <- function(api, sis_id){
  # get species info
  species_info <- #if (FALSE) # don't run
    {assessments_by_sis_id(api,sis_id)}
  list_status <- species_info %>%
    filter(latest == TRUE & scopes_code == 1) #%>%
  #filter(year_published == max(year_published))
  return(list_status$assessment_id)
}
```

```{r}
# Pull iucn_sid
iucn_sid_30 <- aquax_iucn_30$iucn_sid
iucn_sid <- aquax_iucn$iucn_sid
```

```{r}
sample <- assessments_by_sis_id(api, 235)
sample_short <- assessments_by_sis_id(api, 229)
```

```{r}
# Replace incorrect ids (most of these are referencing a subspecies - updated to full species id)
# Replace 21861 id with 21856 (repeated so do twice)
iucn_sid[410] <- 21856
iucn_sid[411] <- 21856
iucn_sid[486] <- 3853
iucn_sid[683] <- 41745
iucn_sid[4015] <- 42838
```

```{r}
# Loop through function, save as list
iucn_id_list <- list()
for(i in 1:nrow(aquax_iucn)){
  test <- iucn_id(api, iucn_sid[i])
  iucn_id_list[[i]] <- test
}
```

```{r}
iucn_id_vector <- unlist(iucn_id_list)
```

```{r}
saveRDS(iucn_id_vector, "iucn_id_vector_041726.rds")
```

```{r}
# Run the assessment numbers collected from the iucn_id function
# Note that 16,429 values went into the function, but api found 16,413 unique assessment IDs
iucn_id_data <- assessment_data_many(api, iucn_id_vector, wait_time = 0.5)
```

```{r}
saveRDS(iucn_id_data, "iucn_id_data_041726.rds")
```

```{r}
# run element names to extract
iucn_status <- extract_element(iucn_id_data, "red_list_category")
```

```{r}
# run element names to extract
sis_taxon_id <- extract_element(iucn_id_data, "sis_taxon_id")
```

```{r}
# run element names to extract
iucn_names <- extract_element(iucn_id_data, "taxon")
```

```{r}
iucn_pop_trend <- extract_element(iucn_id_data, "population_trend")
```

```{r}
iucn_pop_trend_clean <- iucn_pop_trend %>% 
  select(assessment_id, description, code) %>% 
  rename(pop_trend_desc = description,
         pop_trend_code = code)
```

```{r}
iucn_sis_status <- left_join(sis_taxon_id, iucn_status, by = "assessment_id")
iucn_sis_status_names <- left_join(iucn_sis_status, iucn_names,
                                           by = "assessment_id") %>% 
  select(sis_id, assessment_id, scientific_name, description, code,
         kingdom_name:infrarank) %>% 
    rename(iucn_category = description,
           iucn_category_code = code,
           iucn_sid = sis_id)
```

```{r}
#there'll be some entries that don't have trend info (data deficient)
iucn_sis_status_names_trends <- left_join(iucn_sis_status_names, iucn_pop_trend_clean)
```


```{r}
# Replace the same sis_ids 
aquax_iucn_clean <- aquax_iucn

aquax_iucn_clean[410,1] <- 21856
aquax_iucn_clean[411,1] <- 21856
aquax_iucn_clean[486,1] <- 3853
aquax_iucn_clean[683,1] <- 41745
aquax_iucn_clean[4015,1] <- 42838
```

```{r}
iucn_sis_status_names_trends_speciesonly <- iucn_sis_status_names_trends %>% 
  filter(species == TRUE)
```

```{r}
# no multijoins now (subpopulations, infraranks removed)
aquax_iucn_clean_speciesonly <- inner_join(aquax_iucn_clean, iucn_sis_status_names_trends_speciesonly)
```

```{r}
# merge the iucn data with the crosswalk (including subpopulations, etc.)
aquax_iucn_clean_threat <- inner_join(aquax_iucn_clean, iucn_sis_status_names_trends)

```
```{r}
write_csv(aquax_iucn_clean_threat, "/home/abowd/R/repos/aquax_sample/data/iucn_to_worms_match_redlist_category.csv")
```

```{r}
# save to aurora, since more species iucn data than just the 564 AquaX sample
write_csv(aquax_iucn_clean_threat,
          "/home/shares/usgs-cap-rangeshifts/data_iucn/iucn_to_worms_match_redlist_category.csv")
```

```{r}
aquax_iucn_filtered <- aquax_iucn_clean_threat %>% 
  filter(species == TRUE)
```

# Read in sample set of AquaX species

Associate SDM files with the metadata for the 564 sampled species
```{r}
# Not on server
load('data/META_02122025.Rdata')

spp_fs <- list.files('/home/shares/usgs-cap-rangeshifts/data_aquax/2026_OHARA_TST_MORPHO/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

## See ifthe worms name from the AquaX data will match the IUCN x WoRMS crosswalk

```{r}
df_worms <- df %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  rename(spp = species)
```

```{r}
# Joining only the species from IUCN (no subpopulations)
df_worms_iucn <- left_join(df_worms, aquax_iucn_filtered)
```

```{r}
# May be duplicates
# Looks like each of the following WoRMS/AquqX species: "conasprella emarginata" "conus australis"        "jasus paulensis"        "mobula kuhlii" are split into two species each in IUCN
# Might just remove these for now
df_worms_iucn_dupes <- df_worms_iucn %>% 
  get_dupes(worms_name)
```


# Seems like alot of the species that don't match are not evaluated
Using Gemini and fact-checking, looks like almost all of the species without matches are not evaluated by the IUCN. 
Balistoides conspicillum and Pseudobalistes fuscus are listed as LC but weren't on the old crosswalk. Amphiprion latezonatus is DD and not on the old crosswalk. 

```{r}
df_no_iucn <- df_worms_iucn %>% 
  filter(is.na(iucn_sid))
```

```{r}
df_iucn <- df_worms_iucn %>% 
  filter(!is.na(iucn_sid))
```

# Read in EEZ data

```{r}
spp_eez_pxl_count <- read.csv("/home/shares/usgs-cap-rangeshifts/data_aquax/outputs/spp_eez_pxl_count_041626.csv")
```

```{r}
# How many EEZs are each spp present in for current and future?
eez_count_all <- spp_eez_pxl_count %>% 
  group_by(aphiaid) %>% 
  summarize(current_eez_count = sum(n_current > 0),
            RCP85_2100_eez_count = sum(n_RCP85_2100 > 0)) %>% 
  ungroup() #%>% 
  #pivot_longer(cols = c(current, RCP85_2100),
               #names_to = "timeframe", values_to = "eez_count")

```

```{r}
# Green sea turtle drops from 202 EEZs to 2
load("/home/shares/usgs-cap-rangeshifts/data_aquax/2026_OHARA_TST_MORPHO/SDM/FINAL_EMSDM_EMMEAN_SP_137206.Rdata")
```

```{r}
# Black marlin goes from 123 EEZs to 0 Istiompax indica
load("/home/shares/usgs-cap-rangeshifts/data_aquax/2026_OHARA_TST_MORPHO/SDM/FINAL_EMSDM_EMMEAN_SP_712905.Rdata")

# There are no pixels in RCP85_100 that are greater than or equal to cutoff value
```

# Look at EEZs and their count of new/loss species

```{r}
eez_loss_gain <- spp_eez_pxl_count %>% 
  group_by(GEONAME, status) %>% 
  summarize(n = n())
```

```{r}
library(sf)
```

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez_loss_gain_sf <- left_join(eez_loss_gain, eez)
```

```{r}
eez_loss_gain_only_sf <- eez_loss_gain_sf %>% 
  filter(status %in% c("entry into new", "exit from historical"))
```

```{r}
ggplot() +
  geom_sf(data = eez_loss_gain_only_sf, aes(geometry = geom, fill = n)) +
  scale_fill_viridis_c(limits = c(1,51)) +
  theme_bw() +
  facet_wrap(~status) +
  labs(title = "Species count by EEZ (min = 1, max = 47)")
```



```{r}
ggplot() +
  geom_sf(data = eez_loss_gain_sf %>% 
            filter(status == "continued presence"),
          aes(geometry = geom, fill = n)) +
  scale_fill_viridis_c(limits = c(1, 250)) + 
  theme_bw() +
  labs(title = "Species count: continued presence")
```

```{r}
eez_c_f_sf <- eez_loss_gain_sf %>% 
  st_drop_geometry %>% 
  mutate(status = str_replace_all(status, " ", "_")) %>% 
  pivot_wider(names_from = status, values_from = n) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new)
```

```{r}
eez_loss_gain_only <- eez_loss_gain_only_sf %>% 
  mutate(status = str_replace_all(status, " ", "_")) %>% 
  pivot_wider(names_from = status, values_from = n)
```

```{r}
ggplot() +
  geom_point(data = eez_c_f_sf, aes(x = exit_from_historical, y = entry_into_new,
                                            color = Y_1), alpha = 0.5,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Species loss and gain by EEZ",
       y = "species gain count",
       x = "species loss count") +
  theme_bw()
```

```{r}
ggplot() +
  geom_point(data = eez_c_f_sf, aes(x = current_spp, y = future_spp,
                                            color = Y_1), alpha = 0.75) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Current and future (RCP 8.5 2100) by EEZ",
       y = "future species count",
       x = "current species count") +
  theme_bw()
```

## What about only looking at threatened species?
Filter for Vulnerable, Endangered, and Critically Endangered

```{r}
# Only 46 IUCN threatened species 
df_threatened <- df_worms_iucn %>% 
  filter(iucn_category %in% c("Vulnerable", "Endangered", "Critically Endangered")) %>% 
  # drop " Mobula eregoodoo" from scientific name b/c it is an IUCN duplicate of Mobula Kuhlii and is not in AquaX data
  filter(scientific_name != "Mobula eregoodoo")
```

```{r}
threatened_aphiaid <- df_threatened$aphiaid
```

```{r}
# Fiter EEZ pixel data for threatened aphiaid
spp_eez_pxl_count_threat <- spp_eez_pxl_count %>% 
  filter(aphiaid %in% threatened_aphiaid)
```

```{r}
# Count the # of threatened species entering and leaving EEZs
eez_loss_gain_threat <- spp_eez_pxl_count_threat %>% 
  group_by(GEONAME, status) %>% 
  summarize(n = n())
```

```{r}
eez_loss_gain_sf_threat <- left_join(eez_loss_gain_threat, eez)
```

```{r}
eez_loss_gain_only_sf_threat <- eez_loss_gain_sf_threat %>% 
  filter(status %in% c("entry into new", "exit from historical"))
```

```{r}
ggplot() +
  geom_sf(data = eez_loss_gain_only_sf_threat, aes(geometry = geom, fill = n)) +
  scale_fill_viridis_c() +
  theme_bw() +
  facet_wrap(~status) +
  labs(title = "Threatened species count by EEZ")
```

```{r}
ggplot() +
  geom_sf(data = eez_loss_gain_sf_threat %>% 
            filter(status == "continued presence"),
          aes(geometry = geom, fill = n)) +
  scale_fill_viridis_c() + 
  theme_bw() +
  labs(title = "Threatened pecies count: continued presence")
```

```{r}
eez_c_f_sf_threat <- eez_loss_gain_sf_threat %>% 
  st_drop_geometry %>% 
  mutate(status = str_replace_all(status, " ", "_")) %>% 
  pivot_wider(names_from = status, values_from = n) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new)
```

```{r}
ggplot() +
  geom_point(data = eez_c_f_sf_threat, aes(x = exit_from_historical, y = entry_into_new,
                                            color = Y_1), alpha = 0.5,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Threatened species loss and gain by EEZ",
       y = "species gain count",
       x = "species loss count") +
  theme_bw()
```

```{r}
ggplot() +
  geom_point(data = eez_c_f_sf_threat, aes(x = current_spp, y = future_spp,
                                            color = Y_1), alpha = 0.5,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Current and future (RCP 8.5 2100) by EEZ",
       y = "future threatened species count",
       x = "current threatened species count") +
  theme_bw()
```
