---
title: "Regression to predict IUCN status"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(terra)
library(sf)
library(janitor)
```

This document will run a regression to predict AquaX species IUCN status on the number of EEZs its range intersects, whether its range is in the area beyond national jurisdictions (ABNJ), its total range area, its ecoregion, and taxon.

## Prep data 

The number of EEZs and IUCN status is prepared in other files. Here, we need to determine whether each species range intersects with the ABNJ, its total range area, and its ecoregin. The species taxon is included in the AquaX data.

```{r}
# Load ABNJ (high seas)
abnj <-  st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_High_Seas_v2_20241010')
```

```{r}
# Load MEOW (marine ecoregions of the world)
meow <-  st_read('/home/shares/usgs-cap-rangeshifts/spatial/MEOW-TNC')
```

```{r}
ggplot() +
  geom_sf(data = meow, aes(geometry = geometry, fill = REALM))
```

```{r}
meow_realm <- meow %>% 
  group_by(REALM) %>% 
  summarize(geometry = st_combine(geometry)) 
```

```{r}
ggplot() +
  geom_sf(data = meow_realm %>% filter(REALM == "Central Indo-Pacific"),
          aes(geometry = geometry, fill = REALM))
```

```{r}
load('/home/shares/usgs-cap-rangeshifts/data_aquax/2026_OHARA_TST_MORPHO/SDM/FINAL_EMSDM_EMMEAN_SP_550138.Rdata')
```

```{r}
f_test <- FINALEMMEAN %>% 
  select(x, y, Current, cutoff) %>% 
  mutate(historical = case_when(Current < cutoff ~ 0,
                                Current >= cutoff ~ 1)) %>% 
  # Filter for presence
  filter(historical == 1)

f_test_rast <- rast(f_test)
```

```{r}
# Extract maximum value from species range in ABNJ (would be 1 if present in ABNJ)
max_value_hs <- extract(f_test_rast$historical, abnj, fun = max, na.rm = TRUE)
```

```{r}
# Count species pixels in each MEOW realm
meow_count <- extract(f_test_rast$historical, meow_realm, fun = sum, na.rm= TRUE)

colnames(meow_count)[2] <- "presence_count"

meow_count_region <- cbind(meow_realm, meow_count)
```

```{r}
assigned_ecoregion <- meow_count_region[which.max(meow_count_region$presence_count), ]
```

## Build species information

```{r}
# Not on server
load('data/META_02122025.Rdata')
```

Associate SDM files with the metadata for the 564 sampled species
```{r}
spp_fs <- list.files('/home/shares/usgs-cap-rangeshifts/data_aquax/2026_OHARA_TST_MORPHO/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

# Test sample 
A few issues to work out if regression is promising: global species, those species that might not fall into an eco realm, etc.

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df, n = 30)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
group_30 <- test_spp_30$GROUP
#load(test_spp_30$f)

```

```{r}
extract_region <- function(id, df) {
  results_list <- list()
  
  spp_filter <- df %>% 
    filter(aphiaid == id) 
  
  load(spp_filter$f)
  
  spp_presence <- FINALEMMEAN %>% 
  mutate(aphiaid = id) %>% 
  select(x, y, aphiaid, Current, cutoff) %>% 
  mutate(current_presence = case_when(Current < cutoff ~ 0,
                                Current >= cutoff ~ 1)) %>%
  # filter only for presence (1)
  filter(current_presence == 1)
  
  r_spp <- rast(spp_presence, crs = "EPSG:4326")
  
  # high seas flag (abnj)
  hs_extract <- extract(r_spp, abnj, fun = max, na.rm = TRUE)
  # if max is 1 then spp is in high seas
  in_high_seas <- any(hs_extract[, 2] == 1, na.rm = TRUE)
  
  # ecoregion (max pixels in ecoregion)
  realm_extract <- extract(r_spp$current_presence, meow_realm, fun = sum, na.rm = TRUE)
  # get max
  colnames(realm_extract)[2] <- "presence_count"
  realm_count <- cbind(st_drop_geometry(meow_realm), realm_extract)
  assigned_realm <- realm_count[which.max(realm_count$presence_count), 1]
  
  
  results_df <- data.frame(
    aphia_id = id,
    abnj_flag = in_high_seas,
    eco_realm = assigned_realm
  )
  
  #spp_region_df <- rbind(results_df)

  return(results_df)
}
```

```{r}
region_df_30 <- lapply(aphiaid_30, 
                       FUN = extract_region,
                       df = test_spp_30)
```


```{r}
region_df_30_df <- bind_rows(region_df_30)
```

## Run full sample, try parallel

```{r}
aphiaid <- df$aphiaid
```

```{r}
# Time
system.time({
  out_list <- parallel::mclapply(aphiaid, FUN = extract_region, df = df, mc.cores = 10)
})
```

```{r}
out_list_df <- bind_rows(out_list)
```
