---
title: "Traits and AquaX Sample - MPAs"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(sf)
library(terra)
```

# Read in global marine data from Marine Regions
```{r}
marine <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_boundaries_v12.gpkg')
```
```{r}
marine_aus <- marine %>% 
  filter(TERRITORY1 == "Australia")
```


```{r}
plot(marine_aus$geom)
```

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
ggplot() +
  geom_sf(data = eez, fill = "blue")
```

Have EEZ shapefile, but looking for country outlines for mask

```{r}
library(rnaturalearth)
countries <- ne_download(scale = 10, type = "countries")
```

```{r}
plot(countries$geometry)
```


# Read in SSI and MPA data
```{r}
# Species SSI (RCP 8.5, change to 2050)
spp_ssi <- rast("/home/shares/usgs-cap-rangeshifts/data_aquax/outputs/spp_ssi_03.16.26.tif")
```

```{r}
# Check out the MPA file database
st_layers('/home/shares/usgs-cap-rangeshifts/data_wdpa_mpa_2026/WDPA_WDOECM_Mar2026_Public_marine.gdb')
```

```{r}
# Focusing on polygons, not points
mpa_gdb <- vect('/home/shares/usgs-cap-rangeshifts/data_wdpa_mpa_2026/WDPA_WDOECM_Mar2026_Public_marine.gdb',
                   "WDPA_WDOECM_poly_Mar2026_marine")
```

```{r}
# Transform to SSI crs
mpa_gdb_crs <- project(mpa_gdb, crs(spp_ssi))
```

```{r}
# Filter for only "Designated" STATUS
des <- which(mpa_gdb_crs$STATUS == 'Designated')
mpa_gdb_crs_des <- mpa_gdb_crs[des,]
```

```{r}
# Filter one set for no take area greater than or equal to 0.75 and one for categories
ge_75 <- which(mpa_gdb_crs_des$NO_TK_AREA/mpa_gdb_crs_des$REP_M_AREA >= 0.75)
mpa_des_75 <- mpa_gdb_crs_des[ge_75,]

high_cat <- which(mpa_gdb_crs_des$IUCN_CAT %in% c("Ia", "Ib", "II"))
mpa_des_cat <- mpa_gdb_crs_des[high_cat,]
```

```{r}
# Maybe I could do both together - yes!
ge_75_cat <- which(mpa_gdb_crs_des$IUCN_CAT %in% c("Ia", "Ib", "II") |
                 mpa_gdb_crs_des$NO_TK_AREA/mpa_gdb_crs_des$REP_M_AREA >= 0.75)

mpa_ge_75_cat <- mpa_gdb_crs_des[ge_75_cat,]
```

```{r}
# Create a blank raster matching AquaX resolution
mpa_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05)
```

```{r}
mpa_raster_output <- rasterize(mpa_gdb_crs, mpa_raster_blank, field = "IUCN_CAT")
```

```{r}
plot(mpa_raster_output)
```
```{r}
# Create a 10 factor finer blank raster
mpa_raster_blank_005 <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.005)
```

```{r}
mpa_filtered_raster_output <- rasterize(mpa_ge_75_cat, 
                                        mpa_raster_blank_005, 
                                        field = "IUCN_CAT")
```

```{r}
plot(mpa_filtered_raster_output)
```

```{r}
# Assign 1 to the filtered MPAs and 0 to non-MPA areas
binary_mpa <- ifel(is.na(mpa_filtered_raster_output), 0, 1)
```

```{r}
plot(binary_mpa)
```
## Mask land 

```{r}
# Algin CRS of countries
countries_crs <- st_transform(countries, st_crs(spp_ssi))
```

```{r}
# Rasterize countries
countries_r <- rasterize(countries_crs, mpa_raster_blank_005, field = "SOVEREIGNT")
```

```{r}
# Mask land on mpa (inverse = TRUE since keeping NA areas of land raster)
mpa_binary_ocean <- mask(binary_mpa, countries_r, inverse = TRUE)
```

```{r}
plot(mpa_binary_ocean)
```


## Bring in the AquaX data
Bring in the AquaX processed data from traits_aquax_sample.qmd and traits_cont_aquax_sample.qmd scripts. Align with the various MPA IUCN categories spatially. Then, perhaps can move from spatial data to dataframe to calculate statistics of interest

```{r}
plot(spp_ssi)
```
```{r}
# Re-project MPA crs to match SSI crs
mpa_raster <- project(mpa_raster_output, crs(spp_ssi))
```

```{r}
# Create a blank global raster with same resolution and crs; assign pixel ids
id_raster <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05, 
                  crs = crs(spp_ssi))

values(id_raster) <-  1:ncell(id_raster)
names(id_raster) <- "pixel_id"
```

```{r}
plot(id_raster)
```

```{r}
# Extend SSI to full globe extent (id_raster and mpa already match extent)
ssi_raster_ext <- extend(spp_ssi, ext(id_raster))
# Fix name
names(ssi_raster_ext) <- "spp_ssi"
```

```{r}
# Merge rasters
mpa_ssi <- c(id_raster, mpa_raster_output, ssi_raster_ext)
```

```{r}
plot(mpa_ssi)
```
```{r}
# Find the mean species SSI in each IUCN category
zonal(mpa_ssi$spp_ssi, mpa_ssi$IUCN_CAT, "mean")
```
## Read in trait data

```{r}
# Trait SSI (RCP 8.5, change to 2050)
trait_ssi <- rast("/home/shares/usgs-cap-rangeshifts/data_aquax/outputs/trait_ssi_031826.tif")
```

```{r}
# Extend SSI to full globe extent (id_raster and mpa already match extent)
trait_raster_ext <- extend(trait_ssi, ext(id_raster))
# Fix name 
names(trait_raster_ext) <- "ssi_trait"
```

```{r}
# Merge rasters
mpa_ssi_trait <- c(mpa_ssi, trait_raster_ext)
```

```{r}
# Find the mean trait SSI in each IUCN category
zonal(mpa_ssi_trait$ssi_trait, mpa_ssi_trait$IUCN_CAT, "mean")
```
```{r}
# Pixel frequency of the IUCN categories
freq(mpa_ssi$IUCN_CAT)
```

## Test out creating a dataframe 

```{r}
mpa_ssi_trait_df <- as.data.frame(mpa_ssi_trait, xy = TRUE)
```

Before I get to this point, I'm going to have to mask out the land areas for the SSI calculations, otherwise I won't be able to see the difference between land and non-MPA ocean

For, now let's run a t-test comparing the SSI of categories Ia, Ib, & II to everything else

```{r}
mpa_ssi_filter <- mpa_ssi_trait_df %>% 
  dplyr::select(pixel_id, IUCN_CAT, spp_ssi, ssi_trait) %>% 
  filter(IUCN_CAT %in% c("Ia", "Ib", "II", "III", "IV", "V", "VI"))
```

```{r}
mpa_ssi_cat <- mpa_ssi_filter %>% 
  mutate(IUCN_CAT_LVL = case_when(IUCN_CAT %in% c("Ia", "Ib", "II") ~ "high",
         TRUE ~ "low"))
```

```{r}
# t test - species SSI
t.test(spp_ssi ~ IUCN_CAT_LVL, data = mpa_ssi_cat)
```
```{r}
# t test - trait SSI
t.test(ssi_trait ~ IUCN_CAT_LVL, data = mpa_ssi_cat)
```

```{r}
ggplot() +
  geom_density(data = mpa_ssi_cat, aes(x = spp_ssi), fill = "darkgreen", alpha = 0.3) +
  geom_density(data = mpa_ssi_cat, aes(x = ssi_trait), fill = "gray30", alpha = 0.3) +
  facet_wrap(~IUCN_CAT_LVL) +
  labs(x = "SSI")
```

