---
title: "Traits and AquaX Sample"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(terra)
```

```{r}
# Not on server
load('data/META_02122025.Rdata')
```

## Build species information

Associate SDM files with the metadata for the 564 sampled species
```{r}
spp_fs <- list.files('/home/shares/usgs-cap-rangeshifts/data_aquax/2026_OHARA_TST_MORPHO/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

## Inspect a test species

Identify the structure of the SDM .Rdata for a single species

```{r}
# need to figure out this issue 
load('/home/shares/usgs-cap-rangeshifts/data_aquax/2026_OHARA_TST_MORPHO/SDM/FINAL_EMSDM_EMMEAN_SP_550138.Rdata')
```



```{r}
set.seed(123)
test_spp <- slice_sample(df, n = 1)

load(test_spp$f)

summary(FINALEMMEAN)

r <- rast(FINALEMMEAN)
names(r)
plot(r$Current)
plot(r$Current_NR)
plot(r$RCP85_2100)
```

```{r}
# Assign historical, future, (x-y will be pixel id), and change between historical and future 
finalemmean_test <- FINALEMMEAN %>% 
  mutate(aphiaid = test_spp$aphiaid) %>% 
  select(x, y, aphiaid, Current, RCP85_2050, cutoff) %>% 
  mutate(historical = case_when(Current < cutoff ~ 0,
                                TRUE ~ 1),
         future = case_when(RCP85_2050 < cutoff ~ 0,
                             TRUE ~ 1)) %>%
  # historical column plus 2 * future gives 0 = absence, 1 = extinction, 2 = invasion, 3 = presence
  mutate(pxl_change = historical + 2 * future)

```

```{r}
r_pa <- rast(finalemmean_test)
plot(r_pa$historical)
plot(r_pa$future)
plot(r_pa$pxl_change)

```
```{r}
# Invasion and extinction 
finalemmean_inv_ext <- finalemmean_test %>%
  mutate(inv_ext = case_when(historical == 0 & future == 0 ~ 0, #absence
                             historical == 1 & future == 1 ~ 1, #presence
                             historical == 0 & future == 1 ~ 2, #invasion
                             historical == 1 & future == 0 ~ 3,)) #extinction
```

# Test a bigger subsample

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df, n = 30)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
#load(test_spp_30$f)

```

```{r}
test_1 <- test_spp_30 %>% 
  filter(aphiaid == aphiaid_30[18]) 

load(test_1$f)
```


```{r}
extract_change <- function(id, df) {
  spp_filter <- df %>% 
    filter(aphiaid == id) 
  
  load(spp_filter$f)
  
  spp_change <- FINALEMMEAN %>% 
  mutate(aphiaid = id) %>% 
  select(x, y, aphiaid, Current, RCP85_2050, cutoff) %>% 
  mutate(historical = case_when(Current < cutoff ~ 0,
                                TRUE ~ 1),
         future = case_when(RCP85_2050 < cutoff ~ 0,
                             TRUE ~ 1)) %>%
  # historical column plus 2 * future gives 0 = absence, 1 = extinction, 2 = invasion, 3 = presence
  mutate(pxl_change = historical + 2 * future)
  
  r_spp_change <- rast(spp_change)
  
  return(r_spp_change)
}
```

```{r}
change_rast <- lapply(aphiaid_30, FUN = extract_change, df = test_spp_30)
```

```{r}
x <- change_rast[[8]]
```

```{r}
plot(x$pxl_change)
```

## Next, extract the rasters into a dataframe in order to count 
Or stick to raster math (could mask everything but 3s and count those up (the number of species in each cell that remain the same); then count up historical = 1 (historical species) and then could up future = 1 (future species) --> use these to calculate SSI (2 * same / historical + future) 
- challenge - rasters aren't the same size, but cells overlap 

```{r}
raster_x_df <- data.frame(values(x))
```


